---
title: "BST270_Final_Proj"
output: pdf_document
date: "2025-01-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducing "Marriage Isn’t Dead — Yet" Figures

Data are available via on the FiveThirtyEight GitHub (https://github.com/fivethirtyeight/data/tree/master/marriage). First we load the data:

```{r}
# read in data
both_sexes_df <- read.csv("./data/both_sexes.csv")
divorce_df <- read.csv("./data/divorce.csv")
women_df <- read.csv("./data/women.csv")
```

## Figure 1: Marriage Rates by Education and Race - Ages 25-34

We first use the `both_sexes_df`.

Next we simplify dataframes by removing unneeded columns (ie. columns unrelated to education level and race/ethnicity). We also drop na entries and convert divorce rates to marriage rates.

The columns we use are `HS_2534 - High school graduate or less (EDUCD < 65), age between 25 and 34`, `SC_2534 - Some college (EDUCD >= 65 & <= 100), age between 25 and 34`, and `BAp_2534 - Bachelor's degree or more (EDUCD > 100), age between 25 and 34` for the education plot and `White_2534 - Non-Hispanic white, age between 25 and 34`, `Black_2534 - Black or African-American, age between 25 and 34`, and `Hisp_2534 - Hispanic of any race` for the race/ethnicity plot.

```{r}
# Filter both_sexes.csv

# Education
both_sexes_edu_filt_df <- both_sexes_df[, c("year", "HS_2534", 
                                            "SC_2534", "BAp_2534")]
# remove na's from HS_2534, SC_2534, BAp_2534 columns 
both_sexes_edu_filt_df <- both_sexes_edu_filt_df[!is.na(both_sexes_edu_filt_df$HS_2534) | 
                                                 !is.na(both_sexes_edu_filt_df$SC_2534) |
                                                 !is.na(both_sexes_edu_filt_df$SC_2534),]
# convert to marriage rates 
both_sexes_edu_filt_df$HS_2534 <- 1- both_sexes_edu_filt_df$HS_2534 
both_sexes_edu_filt_df$SC_2534 <- 1- both_sexes_edu_filt_df$SC_2534 
both_sexes_edu_filt_df$BAp_2534 <- 1- both_sexes_edu_filt_df$BAp_2534 

# Race & Ethnicity
both_sexes_race_eth_filt_df <- both_sexes_df[, c("year", "White_2534", 
                                                 "Black_2534", "Hisp_2534")]
# remove na's from HS_2534, SC_2534, BAp_2534 columns 
both_sexes_race_eth_filt_df <- both_sexes_race_eth_filt_df[
                                  !is.na(both_sexes_race_eth_filt_df$White_2534) | 
                                  !is.na(both_sexes_race_eth_filt_df$Black_2534) |
                                  !is.na(both_sexes_race_eth_filt_df$Hisp_2534),]
# convert to marriage rates 
both_sexes_race_eth_filt_df$White_2534 <- 1- both_sexes_race_eth_filt_df$White_2534 
both_sexes_race_eth_filt_df$Black_2534 <- 1- both_sexes_race_eth_filt_df$Black_2534 
both_sexes_race_eth_filt_df$Hisp_2534 <- 1- both_sexes_race_eth_filt_df$Hisp_2534 
```

```{r}
# Plot Marriage Rates for Education Level
matplot(both_sexes_edu_filt_df$year, cbind(both_sexes_edu_filt_df$HS_2534, 
                                           both_sexes_edu_filt_df$SC_2534, 
                                           both_sexes_edu_filt_df$BAp_2534), 
        type = "l", lty = 1,lwd=3, 
        col = c("maroon", "magenta", "pink"), xlab = "Year", 
        ylab = "Marriage Rate", main = "Marriage Rates by Education",
        ylim = c(0,1))
grid(nx = NULL, ny = NULL,
     lty = 1,
     col = "gray", 
     lwd = 0.5)
legend("topright", legend = c("High School or Less", "Some College", 
                              "College Graduates"), 
       col = c("maroon", "magenta", "pink"), 
       lty = 1, lwd=3)
```
```{r}
matplot(both_sexes_race_eth_filt_df$year, cbind(both_sexes_race_eth_filt_df$White_2534,
                                           both_sexes_race_eth_filt_df$Black_2534, 
                                           both_sexes_race_eth_filt_df$Hisp_2534), 
        type = "l", lty = 1,lwd=3, 
        col = c("forestgreen", "darkgreen", "lightgreen"), xlab = "Year", 
        ylab = "Marriage Rate", main = "Marriage Rates by Race",
        ylim = c(0,1))
grid(nx = NULL, ny = NULL,
     lty = 1, 
     col = "gray", 
     lwd = 0.5) 
legend("topright", legend = c("Non-hispanic White", "Black", "Hispanic"), 
       col = c("forestgreen", "darkgreen", "lightgreen"), 
       lty = 1, lwd=3)
```

## Figure 3: Marriage Rates for Women by Education - Ages 25-34
RESTART HERE 
We first use the `women_df`.

The columns we use are `kids_HS_2534 - High school graduate or less (EDUCD < 65), age between 25 and 34`, `SC_2534 - Some college (EDUCD >= 65 & <= 100), age between 25 and 34`, and `BAp_2534 - Bachelor's degree or more (EDUCD > 100), age between 25 and 34` for the education plot and `White_2534 - Non-Hispanic white, age between 25 and 34`, `Black_2534 - Black or African-American, age between 25 and 34`, and `Hisp_2534 - Hispanic of any race` for the race/ethnicity plot.


```{r}
# Filter women.csv

# With Children
women_df_with_children <- women_df[, c("year", "kids_HS_2534", "kids_BAp_2534")]

# remove na's from kids_HS_2534 and kids_BAp_2534 columns 
women_df_with_children <- women_df_with_children[
                                    !is.na(women_df_with_children$kids_HS_2534) | 
                                    !is.na(women_df_with_children$kids_BAp_2534),]
# convert to marriage rates
women_df_with_children$kids_HS_2534 <- 1- women_df_with_children$kids_HS_2534
women_df_with_children$kids_BAp_2534 <- 1 - women_df_with_children$kids_BAp_2534

# Without Children
women_df_without_children <- women_df[, c("year", "nokids_HS_2534", "nokids_BAp_2534")]

# remove na's from nokids_HS_2534 and nokids_BAp_2534 columns 
women_df_without_children <- women_df_without_children[
                              !is.na(women_df_without_children$nokids_HS_2534) | 
                              !is.na(women_df_without_children$nokids_BAp_2534),]
# convert to marriage rates
women_df_without_children$nokids_HS_2534 <- 1- women_df_without_children$nokids_HS_2534
women_df_without_children$nokids_BAp_2534 <- 1 - women_df_without_children$nokids_BAp_2534
```


```{r}
# With Children
matplot(women_df_with_children$year, cbind(women_df_with_children$kids_HS_2534, 
                                           women_df_with_children$kids_BAp_2534), 
        type = "l", lty = 1,lwd=3, 
        col = c("plum", "purple"), xlab = "Year", 
        ylab = "Marriage Rate", main = "Marriage Rates for Women - With Children",
        ylim = c(0,1))
grid(nx = NULL, ny = NULL,
     lty = 1,
     col = "gray", 
     lwd = 0.5)
legend("bottomright", legend = c("High School or Less",
                              "College Graduates"), 
       col = c("plum", "purple"), 
       lty = 1, lwd=3)
```

```{r}
# Without Children
matplot(women_df_without_children$year, cbind(women_df_without_children$nokids_HS_2534, 
                                           women_df_without_children$nokids_BAp_2534), 
        type = "l", lty = 1,lwd=3, 
        col = c("plum", "purple"), xlab = "Year", 
        ylab = "Marriage Rate", main = "Marriage Rates for Women - Without Children",
        ylim = c(0,1))
grid(nx = NULL, ny = NULL,
     lty = 1,
     col = "gray", 
     lwd = 0.5)
legend("bottomright", legend = c("High School or Less",
                              "College Graduates"), 
       col = c("plum", "purple"), 
       lty = 1, lwd=3)
```

## Figure 4:

### Divorce Rates by Education

```{r}
# Divorce rates
divorce_edu_df <- divorce_df[,c("year", "HS_3544", "SC_3544", "BAp_3544")]
divorce_edu_df <- divorce_edu_df[
                      !is.na(divorce_edu_df$HS_3544) | 
                      !is.na(divorce_edu_df$SC_3544) | 
                      !is.na(divorce_edu_df$BAp_3544) ,]
```

```{r}
# Divorce Plot
matplot(divorce_edu_df$year, cbind(divorce_edu_df$HS_3544, 
                                   divorce_edu_df$SC_3544,
                                   divorce_edu_df$BAp_3544), 
        type = "l", lty = 1,lwd=3, 
        col = c("turquoise", "blue", "lightblue"), xlab = "Year", 
        ylab = "Divorce Rate", main = "Divorce Rates by Education - Ages 35-44",
        ylim = c(0,0.25))
grid(nx = NULL, ny = NULL,
     lty = 1,
     col = "gray", 
     lwd = 0.5)
legend("topleft", legend = c("High School or Less",
                                 "Some College",
                                 "College Graduates"), 
       col = c("turquoise", "blue", "lightblue"), 
       lty = 1, lwd=3)
```
